---
title: "Multiple and spatial regression analysis - Assignment 5"
author: "Edgar Manrique"
date: "5/17/2022"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: true
---

We are going to explore data from the London Ward Stats, trying to fit a multivariate linear model to predict Average GCSE scores. As an example we are going to use the median age of the ward, continuing the work from the last assignment.

```{r, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(patchwork)
library(RColorBrewer)

df <- st_read('~/Desktop/SpatStat/class 5/LondonWardStats/LondonWardStats.shp')
Regressiondata <- df %>%
  dplyr::select(BOROUGH, Average.GC, Median.Age, Median.H_1, Crime.rate, Employment, Unauthoris, Median.Hou) |> 
  filter(complete.cases(Average.GC, Median.Age, Median.H_1, Crime.rate, Employment))
```

We are adding some additional variables to help improve the fitness of our model. 

```{r}
Regressiondata |> as.data.frame() |> select(-geometry, -BOROUGH) |> 
  gather(key = 'parameter', value = 'value') |> 
ggplot() + 
  geom_histogram(aes(x = value)) + 
  facet_wrap(.~parameter, scales = 'free')+
  theme_minimal()
```

We can see that some of the variables do not have a normal distribution. We can fix this by transforming the values, for example using a natural logarithm transformation.

```{r}
Regressiondata |> as.data.frame() |> select(-geometry, -BOROUGH) |> 
  gather(key = 'parameter', value = 'value') |> 
  mutate(value = if_else(parameter %in% c('Median.H_1', 'Crime.rate', 'Median.Hou'), 
                         log(value),
                         value)) |> 
ggplot() + 
  geom_histogram(aes(x = value)) + 
  facet_wrap(.~parameter, scales = 'free')+
  theme_minimal()
```

Now all possible predictors follow a somewhat normal distribution.

We can also explore co-linearity of the variables with correlation

```{r}
library(GGally)
Regressiondata |> 
  as.data.frame() |> 
  select(-geometry, -BOROUGH) |> 
  mutate(Median.H_1 = log(Median.H_1),
         Crime.rate = log(Crime.rate),
         Median.Hou = log(Median.Hou)) |> 
  ggpairs()
```

In this case Median.H_1 and Median.Hou, and Median.H_1 and Employment have a high positive correlation, so it would not be recommended to use those variables together.

We are going to add variables to our previous linear model and see which combination explains better the variability of Average GCSE scores.

```{r}
#now model
model1 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age, data=.)
model2 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + log(Median.H_1), data=.)
model3 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + log(Crime.rate), data=.)
model4 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + Employment, data=.)
model5 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + Unauthoris, data=.)
model6 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + log(Median.Hou), data=.)
model7 <- Regressiondata %>%
  lm(Average.GC ~ Median.Age + log(Median.H_1) + Unauthoris, data=.)

#show the summary of those outputs
summary(model1)
summary(model2)
summary(model3)
summary(model4)
summary(model5)
summary(model6)
summary(model7)

```

Median.Age, Median.H_1 and Unauthoris explain 54% of Average GCSE scores variability. 

```{r}
Regressiondata <- Regressiondata |> 
  mutate(residuals = residuals(model7))
```

```{r}
(g1 <- ggplot(Regressiondata) + 
  geom_sf(aes(fill = residuals), color = 'white', size = .2) +
  scale_fill_distiller(
  type = "div",
  palette = 5,
  direction = 1,
  aesthetics = "fill"
) +
  coord_sf() +
  theme_minimal() +
  labs(fill = 'Model residuals')
)
```


```{r, message=FALSE, warning=FALSE}
library(spatialreg)
library(spdep)

knn <- st_geometry(Regressiondata) %>%
  st_centroid() %>%
  knearneigh(., k=4) %>%
  knn2nb()

lw <- nb2listw(knn, style="W", zero.policy=TRUE)
slag_model2_queen <- lagsarlm(Average.GC ~ Median.Age + log(Median.H_1) + Unauthoris,
                              data = Regressiondata,
                              lw, method = "eigen")
```


```{r, message=FALSE, warning=FALSE}
moran.mc(Regressiondata$residuals, lw, nsim=9999, alternative="greater")
```

Moran's Index was calculated to test for spatial autocorrelation of the residuals. Now we are going to introduce a spatial lag term to the equation.

```{r, message=FALSE, warning=FALSE}
broom::tidy(slag_model2_queen)
summary(slag_model2_queen)

Regressiondata <- Regressiondata |> 
  mutate(residuals_spatial = residuals(slag_model2_queen))

moran.mc(Regressiondata$residuals_spatial, lw, nsim=9999, alternative="greater")
```

By compensating for the spatial variability, Moran's I was reduced significantly. 

```{r}
g2 <- ggplot(Regressiondata) + 
  geom_sf(aes(fill = residuals_spatial), color = 'white', size = .2) +
  scale_fill_distiller(
  type = "div",
  palette = 5,
  direction = 1,
  aesthetics = "fill"
) +
  coord_sf() +
  theme_minimal() +
  labs(fill = 'Model residuals - spatial')

g1 / g2
```

```{r}
library(ggExtra)

g <- Regressiondata |> 
  ggplot() + 
  geom_point(aes(x = residuals, y = residuals_spatial)) +
  xlim(-50, 50) +
  ylim(-50, 50) +
  coord_equal()

g1 <- ggMarginal(g, type="density")
g1
```

Model residuals for both, multiple linear regression with and without the spatial term, have a normal distribution. 

