---
title: "Assignment 6 - Spatial autocorrelation"
author: "Edgar Manrique"
date: "5/31/2022"
output: html_document
---

```{r, message=FALSE}
library(tidyverse)
```

# Spatial autocorrelation

```{r}
m <- c(0, 1, 1, 0, 0, 0,
       1, 0, 1, 1, 1, 0,
       1, 1, 0, 0, 1, 1, 
       0, 1, 0, 0, 1, 0, 
       0, 1, 1, 1, 0, 1, 
       0, 0, 1, 0, 1, 0) 
dim(m) <- c(6, 6)

v <- c(32, 26, 19, 18, 17, 14,
       32, 26, 19, 18, 17, 14,
       32, 26, 19, 18, 17, 14,
       32, 26, 19, 18, 17, 14,
       32, 26, 19, 18, 17, 14,
       32, 26, 19, 18, 17, 14)
dim(v) <- c(6, 6)

v_t <- t(v)

mean <- mean(v)

u <- m*(v-mean)*(v_t-mean)

numerator <- sum(u)
sum_weights <- sum(m)
n <- 6
sum_squares <- sum((v_t[ 1, ] - mean)^2)

denominator <- sum_squares/(n-1)*sum_weights

moranI <- numerator/denominator
```

```{r}
m_list <- spdep::mat2listw(m, style = 'B')

spdep::moran.mc(v_t[ 1, ], m_list, nsim=500)
```

## What is the observed rank of the observed Moran's I?

-   The observed rank is 499.

## What is p and how is it calculated?

-   The p value is 0.003992, it is calculated from the number of simulations.

## Formulate H0 and H1. Is your result significant? On what significance level? What does that mean in terms of H0 and H1?

-   Null hypothesis = No difference between the observed and the expected autocorrelation.

-   Alternative hypothesis = The observed autocorrelation is greater then the expected autocorrelation.

-   The p-value is significant \< 0.05. The null hypothesis is rejected.
